---
title: "Wisconsin's Tax System and the Property-Tax Landscape"
subtitle: "Part I of the internal technical companion to the policy brief — descriptive baseline: who taxes, what they may levy, what they collect, what they spend, and the household income-tax baseline. Part II, the replacement model, is tax_scenarios_model.qmd."
author: "Alexander J. Ryan"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
    fig-dpi: 300
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "png"
    echo: false
---

```{=html}
<!-- INTERNAL WORKING DOCUMENT (Part I of two) -- Lade lab / Swank Program.
     Companion to tax_scenarios_model.qmd (Part II, the replacement model).
     This document is the descriptive baseline: who taxes, what they may
     levy, what they collect, what they spend, and the income-tax baseline
     that Part II's income-replacement scenarios build on. Every number
     below is READ from a file the Makefile already builds -- an inline R
     read of a pipeline output, or a labeled external statutory/historical
     fact -- nothing here is recomputed. If a number in this document ever
     disagrees with the file it cites, the file is right and this document
     is stale; fix the document, never adjust a script to match this one.
     Structure: (i) who are the taxing authorities, (ii) what each may levy
     and at what rates, (iii) what they collect, (iv) what they spend --
     then the ACS/PUMS incidence baseline (property tax and income tax) and
     what's next. Code is hidden (echo:false); every figure cites its
     source file in its caption. Renders top-to-bottom (confirmed Codex QC
     2026-06-19; re-confirmed 2026-07-16); re-run after any data refresh. -->
```

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(sf)
library(scales)
library(patchwork)
library(arrow)
library(leaflet)
library(here)

source(here("scripts/functions/theme_swank.R"))     # theme_swank(), swank_palette
# source(here("02_code/functions/read_odt.R"))         # read_odt(), keep_oh_counties()
source(here("scripts/functions/viz_helpers.R"))      # theme_swank_map(), maps, scales
source(here("scripts/functions/incidence.R"))        # property_tax_incidence(), wtd_quantile_bins()

fig_dir <- here("output/figures")

# Geographies, reprojected to EPSG:3735 (area-correct Ohio). County cb file is
# shoreline-clipped, so Lake Erie is not a polygon.
wi_shp  <- st_read("C:/Users/aryan/Documents/GitHub/Wisconsin_Tax/data/shapefiles/CTV_Jan_2026_-4910906020273727621/CTV_Jan_2026.shp")

wi_city <- wi_shp %>% filter(CTV == "C") %>% st_transform(3735)
wi_village <- wi_shp %>% filter(CTV == "V") %>% st_transform(3735)
wi_township <- wi_shp %>% filter(CTV == "T") %>% st_transform(3735)
wi_county <- st_read("C:/Users/aryan/Documents/GitHub/Wisconsin_Tax/data/shapefiles/cb_2018_us_county_500k/cb_2018_us_county_500k.shp") %>% filter(STATEFP == 55) %>% st_transform(3735)

wi_sd        <- st_read("C:/Users/aryan/Documents/GitHub/Wisconsin_Tax/data/shapefiles/Wisconsin_School_Districts_-5132457395192808409/WI_School_Districts.shp") |> st_transform(3735)

# Wisconsin land polygon (counties are shoreline-clipped, so their union excludes Lake
# Erie). Clip the other jurisdiction layers to it so the lake doesn't render.
wi_land        <- sf::st_union(sf::st_make_valid(wi_county))
wi_cities_l    <- clip_to_land(wi_city, wi_land)
wi_townships_l <- clip_to_land(wi_township, wi_land)
wi_villages_l  <- clip_to_land(wi_village, wi_land)
wi_sd_l        <- clip_to_land(wi_sd, wi_land)
n_cities <- nrow(wi_city); n_twp <- nrow(wi_township); n_sd <- nrow(wi_sd)
n_villages <- nrow(wi_village)

```

This document builds the descriptive baseline that the replacement-scenario model (`02_code/scenarios/`, Part II) perturbs when estimating the consequences of abolishing Ohio's **real property tax**. We organize it around four questions — *who* taxes, *what* they may levy, *what* they collect, and *what* they spend — then turn to how the ACS/PUMS household data converts those fiscal totals into a household-incidence baseline, for both the property tax and the income tax. The recurring finding: property tax is a **local, school-centred, real-property** instrument distributed unevenly across Ohio, so a statewide replacement is never a clean swap.

# (i) Who are the taxing authorities?

Ohio property tax is levied not by one government but by a stack of overlapping ones — counties, municipalities (cities and villages), townships, and school districts, plus special districts (libraries, parks, transit). A single parcel typically pays into several at once. That overlap is why "abolish the property tax" is hundreds of simultaneous fiscal events, not one.

```{r}
# #| label: fig-jurisdictions
# #| fig-cap: "Ohio's four overlapping kinds of property-taxing authority (Lake Erie removed)."
# #| fig-width: 10
# #| fig-height: 8
# base_map <- function(layer, title)
#   ggplot() +
#     geom_sf(data = layer, fill = swank_palette[1], color = "white", linewidth = 0.08, alpha = 0.85) +
#     geom_sf(data = wi_land, fill = NA, color = "grey30", linewidth = 0.3) +
#     labs(title = title) + theme_swank_map(base_size = 11)

# p_jur <- (base_map(wi_county, "Counties") | base_map(wi_cities_l, paste(n_cities, "Municipalities"))) /
#          base_map(wi_villages_l, paste(n_villages, "Villages")) | (base_map(wi_townships_l, paste(n_twp, "Townships")) | base_map(wi_sd_l, paste(n_sd, "School districts"))) +
#   plot_annotation(
#     title = "Four overlapping kinds of taxing authority levy Ohio's property tax",
#     subtitle = "Each parcel sits inside a county, a municipality or township, and a school district at once",
#     caption = "Source: Census TIGER (counties, school districts) + ODOT/OGRIP (Ohio_City_Boundaries.shp, Ohio_Township_Boundaries.shp).",
#     theme = theme(plot.title = element_text(face = "bold", size = 15))
#   )
# p_jur
# ggsave(file.path(fig_dir, "fig04_jurisdictions_small_multiples.png"),
#        p_jur, width = 10, height = 8, dpi = 300, bg = "white")
```

School districts and townships tile the whole state; municipalities are islands of denser settlement. The model carries all four levels because replacement revenue routed to one level does not automatically reach the level that lost it — the heart of the Layer-2 redistribution problem.

# (ii) What may each authority levy — and at what rates?

Not every government can levy every tax. This is the single most useful fact for reasoning about replacement: when the property tax disappears, only certain levels have other instruments to fall back on, and the property tax is the *only* broad tax a township or special district has at all.

```{r}
# #| label: prep-sd-income-count
# # Hoisted here (ahead of the section-iii map that otherwise builds it first)
# # so the "how many districts levy a school income tax" figure quoted in
# # this section and in the open item further below is a live count, not a
# # hand-typed one -- same construction the fig-sd-income chunk in section
# # iii uses (ODT Table SD-2, FY2026 Q4); that chunk reuses this object
# # rather than re-deriving it.
# sd2_m <- read_excel(raw("Local Tax/School_Income_Tax/sd2q4fy26.xlsx"),
#                     sheet = "SD2Q4_FY26", col_names = FALSE)
# sd_income <- tibble(
#   irn = as.character(sd2_m[[4]]),                              # col 4 = IRN
#   q1 = suppressWarnings(as.numeric(sd2_m[[9]])),  q2 = suppressWarnings(as.numeric(sd2_m[[10]])),
#   q3 = suppressWarnings(as.numeric(sd2_m[[11]])), q4 = suppressWarnings(as.numeric(sd2_m[[12]]))  # cols 9-12 = quarters
# ) |>
#   filter(!is.na(suppressWarnings(as.numeric(irn)))) |>
#   mutate(sd_income = rowSums(across(c(q1, q2, q3, q4)), na.rm = TRUE)) |>
#   select(irn, sd_income)
# n_sd_income_positive <- sum(sd_income$sd_income > 0)
```

| Authority | Property tax | Sales / use tax | Income tax | Other notable |
|---|:---:|:---:|:---:|---|
| **State of Wisconsin** | — | **✓** 5.0% | **✓** 3.50–7.65% individual | corporate franchise/income, motor fuel, alcohol/tobacco and other excises |
| **County** | **✓** | **✓** generally 0.5%; Milwaukee Co. 0.9% | — | qualifying counties may impose premier resort area tax |
| **City** | **✓** | Limited[^wi-muni-sales] | — | municipal room tax; TIF; special charges/fees |
| **Village** | **✓** | Limited[^wi-muni-sales] | — | municipal room tax; TIF; special charges/fees |
| **Town** | **✓** | Limited[^wi-muni-sales] | — | municipal room tax; TIF; special charges/fees |
| **School district** | **✓** | — | — | referendum-approved levy/debt authority |
| **Technical college district** | **✓** | — | — | — |
| **Special district** (sanitary, sewerage, lake, etc.) | **✓** in many cases | Limited[^wi-special-sales] | — | special assessments / charges |

[^wi-muni-sales]: Municipal sales taxation is very limited in Wisconsin. The **City of Milwaukee** is the only municipality with a general city sales/use tax, at 2%. Qualifying cities, villages, and towns may instead impose a **Premier Resort Area Tax** on specified taxable sales. This authority applies only to designated resort municipalities and is not general municipal sales-tax authority.

[^wi-special-sales]: Special-district sales taxes are unusual rather than a general feature of Wisconsin local taxation. Milwaukee-area local exposition taxes are a notable example and apply to selected activities such as lodging, food and beverages, and rental cars.

: Which Ohio governments may levy which broad taxes

**Notes.** ^†^ *Selective excises* are narrow taxes on specific goods (cigarette, alcohol, motor fuel, etc.) — distinct from the broad **sales / use tax** on general retail, which is the only thing meant by "sales tax" in this column. ^‡^ *Permissive* sales tax: a **county** may levy up to **1.5%**, and a transit authority additional increments; combined local rates reach \~2.25%. ^§^ Townships have **no income-tax authority of their own**. They can share in an income tax only through a **JEDD** (Joint Economic Development District) — a contract with a municipality that imposes the *municipality's* income tax inside a defined zone. **SD** = school district. The **state income tax** is a flat 2.75%, but only on Ohio taxable income **above ≈ \$26,050** — income at or below that threshold owes no state income tax (so the first \~\$26k of taxable income is effectively exempt).


The *rates* themselves vary enormously across space. The map below summarizes school-district property-tax millage using the Wisconsin TVC Tableau extract and the Wisconsin school-district boundaries.

```{r}
#| label: fig-sd-rate
#| fig-cap: "School-district property-tax millage (Mill Rate) from the Wisconsin School District Income and Property Tax Tableau extract, latest available IPAS year."
sd_rate_file_candidates <- c(
  here("data/raw/sd_taxes/school_district_mill_rate_latest.csv"),
  here("scripts/data/raw/sd_taxes/school_district_mill_rate_latest.csv")
)
sd_rate_file <- sd_rate_file_candidates[file.exists(sd_rate_file_candidates)][1]
if (is.na(sd_rate_file)) {
  stop("Missing school_district_mill_rate_latest.csv. Run the Tableau extract pull first.")
}

sd_rates <- readr::read_csv(sd_rate_file, show_col_types = FALSE) |>
  transmute(
    SDID = stringr::str_pad(as.character(district_number), width = 4, side = "left", pad = "0"),
    district_name = as.character(district_name),
    ipas_year = suppressWarnings(as.integer(ipas_year)),
    mill_rate = suppressWarnings(as.numeric(mill_rate))
  ) |>
  filter(!is.na(SDID), !is.na(mill_rate))

latest_rate_year <- max(sd_rates$ipas_year, na.rm = TRUE)

sd_rate_map <- wi_sd_l |>
  st_make_valid() |>
  select(SDID, DISTRICT, geometry) |>
  left_join(sd_rates, by = "SDID")

n_sd_rated <- sum(!is.na(sd_rate_map$mill_rate))
n_sd_total <- nrow(sd_rate_map)

p_fig8 <- ggplot(sd_rate_map) +
  geom_sf(aes(fill = mill_rate), color = "white", linewidth = 0.05) +
  geom_sf(data = wi_land, fill = NA, color = "grey30", linewidth = 0.3) +
  scale_fill_swank_seq(name = "Mill rate (mills)", labels = label_number(accuracy = 0.1)) +
  labs(
    title = "School-district property-tax millage varies across Wisconsin",
    subtitle = paste0("Direct district-level Mill Rate from School District Income and Property Tax Tableau extract (IPAS ", latest_rate_year, ")"),
    caption = paste0(
      "Source: Wisconsin School District Income and Property Tax Tableau workbook + Wisconsin school-district shapefile.\n",
      "Mapped directly by district code (District Number -> SDID), no municipal-weighted overlay."
    )
  ) +
  theme_swank_map()
p_fig8
ggsave(file.path(fig_dir, "fig08_school_district_class1_rate_2025.png"),
       p_fig8, width = 8, height = 7, dpi = 300, bg = "white")
```

Using the latest available IPAS year, `r n_sd_rated` of `r n_sd_total` Wisconsin school districts have a directly mapped district-level mill-rate value.

> **Note.** This version maps district-level values directly from the school-district Tableau extract (no area-weighted municipal interpolation).

# (iii) What does each level collect?

## The macro picture

```{r}
#| label: fig-collections-stacked
#| fig-cap: "Wisconsin state and local tax collections by type, stacked by level (mixed vintages: local CY2024 from TVC Tableau; state FY2025 from WI State and Local Taxes Tableau)."
collections_wi <- readr::read_csv(
  here("data/intermediate/fig3_wi_collections_by_type.csv"), show_col_types = FALSE
) |>
  mutate(collections_b = value_usd / 1e9,
         level = factor(level, levels = c("Local", "State")))

tax_cols_wi <- c(
  "K-12 School Tax" = "#1F3864",
  "Municipal Tax" = "#2E7D32",
  "County Tax" = "#5B9BD5",
  "Technical College Tax" = "#70AD47",
  "Special Districts Tax" = "#A9D18E",
  "TID Tax" = "#7F7F7F",
  "Individual Income" = "#2F5597",
  "State Sales And Use" = "#ED7D31",
  "Corporate Income And Franchise" = "#9E480E",
  "Motor Fuels Tax" = "#A6611A",
  "Utility" = "#8064A2",
  "Cigarette" = "#C0504D",
  "Insurance" = "#4BACC6",
  "Real Estate Transfer Fee" = "#4F6228",
  "Tobacco Products" = "#B5651D",
  "Liquor And Wine" = "#7030A0",
  "Other /Remaining Segregated Fund Taxes" = "#7F7F7F",
  "Fire Dues" = "#948A54",
  "Transportation Fund Taxes (Other Than Motor Fuel)" = "#00B0F0",
  "Lawsuits" = "#A5A5A5",
  "Beer" = "#FFC000",
  "Vapor Products" = "#92D050",
  "County Sales Administrative Fee" = "#BFBFBF",
  "Business Trust Regulation Fee" = "#D99694",
  "Local Expo Administrative Fee" = "#8DB4E2",
  "Other Program Tax Revenues" = "#DDD9C3",
  "Premier Resort Administrative Fee" = "#FABF8F",
  "Other" = "#E6E6E6"
)
ord_wi <- collections_wi |> count(tax_type, wt = collections_b, sort = TRUE) |> pull(tax_type)
stack_df_wi <- collections_wi |> mutate(tax_type = factor(tax_type, levels = rev(ord_wi)))

p_fig3 <- ggplot(stack_df_wi, aes(level, collections_b, fill = tax_type)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(collections_b >= 2,
          paste0(str_wrap(tax_type, 16), "\n$", formatC(collections_b, 1, format = "f"), "B"), "")),
    position = position_stack(vjust = 0.5), size = 2.6, lineheight = 0.85,
    color = "white", fontface = "bold") +
  scale_fill_manual(values = tax_cols_wi, breaks = ord_wi, name = NULL) +
  scale_y_continuous(labels = label_dollar(suffix = "B"), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Wisconsin local tax base is property-tax heavy; state revenue is income and sales-tax heavy",
       subtitle = "Local from TVC district/municipal/county components (CY2024); State from WI State and Local Taxes dashboard (FY2025)",
       x = NULL, y = "Collections",
       caption = "Source: TVC Tableau exports (local) and WIStateandLocalTaxes Tableau workbook (state). Mixed vintages by source.") +
  theme_swank() +
  theme(legend.position = "right", legend.text = element_text(size = 8), panel.grid.major.x = element_blank())
p_fig3
ggsave(file.path(fig_dir, "fig03_collections_state_local_stacked.png"),
       p_fig3, width = 9, height = 6, dpi = 300, bg = "white")
```

```{r}
#| label: fig-collections-ranked
#| fig-cap: "Wisconsin state and local tax sources, ranked by collections (local grouped by tax type family)."
local_family_map <- readr::read_csv(
  here("data/intermediate/fig_local_tax_type_detail_latest.csv"), show_col_types = FALSE
) |>
  transmute(tax_type = tax, tax_type_plot = tax_family)

ranked <- collections_wi |>
  mutate(tax_type_plot = tax_type) |>
  left_join(local_family_map, by = "tax_type", suffix = c("", "_local")) |>
  mutate(tax_type_plot = if_else(level == "Local" & !is.na(tax_type_plot_local), tax_type_plot_local, tax_type_plot)) |>
  group_by(level, tax_type_plot) |>
  summarise(collections_b = sum(collections_b, na.rm = TRUE), .groups = "drop") |>
  mutate(lab = paste0(tax_type_plot, "  (", level, ")")) |>
  arrange(collections_b) |>
  mutate(lab = factor(lab, levels = lab))

p_fig2 <- ggplot(ranked, aes(collections_b, lab, fill = level)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = paste0("$", formatC(collections_b, 1, format = "f"), "B")),
            hjust = -0.12, size = 3, color = "#404040") +
  scale_fill_manual(values = c(State = "#5B9BD5", Local = "#1F3864"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14))) +
  labs(title = "Wisconsin tax sources ranked by collections",
       x = NULL, y = NULL,
       caption = str_wrap(paste(
         "Source: TVC Tableau exports (local, CY2024) and WIStateandLocalTaxes Tableau workbook (state, FY2025).",
         "Local rows are grouped to the same tax-type families used in fig_local_funding_by_tax_type."), 90)) +
  theme_swank() +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        axis.text.x = element_blank(), panel.grid.major.x = element_blank())
p_fig2
ggsave(file.path(fig_dir, "fig02_collections_ranked_bar.png"),
       p_fig2, width = 8, height = 5, dpi = 300, bg = "white")
```


```{r}
#| label: fig-jurisdiction-shares
#| fig-cap: "Share of Wisconsin local property-tax levy by jurisdiction type (FY2025)."
jur <- readr::read_csv(here("data/intermediate/fig12_wi_property_tax_by_jurisdiction.csv"),
                       show_col_types = FALSE) |>
  mutate(ptax = value_m * 1e6) |>
  group_by(jurisdiction_type) |>
  summarise(ptax = sum(ptax, na.rm = TRUE), .groups = "drop") |>
  mutate(pct = round(ptax / sum(ptax, na.rm = TRUE) * 100, 1)) |>
  arrange(pct) |>
  mutate(jurisdiction_type = factor(jurisdiction_type, levels = jurisdiction_type),
         highlight = as.character(jurisdiction_type) == "School District")

p_fig12 <- ggplot(jur, aes(pct, jurisdiction_type, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.15, size = 3.6, color = "#404040") +
  scale_fill_manual(values = c(`TRUE` = swank_palette[2], `FALSE` = "#BFBFBF"), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), limits = c(0, 75)) +
  labs(title = "School districts collect the largest share of Wisconsin local property tax",
       subtitle = "Local gross property-tax levy by jurisdiction type (FY2025)",
       x = NULL, y = NULL,
       caption = "Source: WIStateandLocalTaxes Tableau workbook, local gross levy components (FY2025).") +
  theme_swank() +
  theme(panel.grid.major = element_blank(), axis.text.x = element_blank())
p_fig12
ggsave(file.path(fig_dir, "fig12_property_tax_by_jurisdiction.png"),
       p_fig12, width = 8, height = 4.5, dpi = 300, bg = "white")
```

<!-- This is now **computed from the Government Finance Database**, not lifted from the deck: school districts collect \~65% of Ohio property tax, counties \~17%, townships \~7%, municipalities \~6%, special districts \~4% (JVSDs fall inside "School District" here). The numbers sit close to Alex's deck figures; the publishable version still reconciles to ODT's property-tax tables, which are the statutory source of record. -->

How dependent are schools on the property tax specifically? Of the **own-source (local) revenue** a district raises itself — setting aside state and federal aid — property tax is the overwhelming majority: districts levy the voted school income tax in only a minority of cases (the SD income-tax map further below in this section), so for most districts property tax *is* own-source revenue.

The precise own-source share is not yet computed from a reconciled district-finance source and is flagged here as an open item rather than asserted with false precision — see the note below and next-steps item 2. Two related, already-computed figures bound it from different directions: school districts collect the largest share of statewide property tax among jurisdiction types (the fig12 jurisdiction-type shares, above), and school districts are one of the two most property-tax-reliant levels of government by *total* revenue (§iv below) — a lower bound on the own-source share, since state and federal aid dilute the total-revenue denominator but not the own-source one.


## Collections over time (2005–2023)

Everything above is a single-year cross-section. The Government Finance Database (GFD) covers Ohio annually back to 2005 (full-coverage census years 2007/2012/2017/2022, an annual-survey sample in between), so we can ask whether property tax's role has been shrinking or growing relative to the other major instruments — directly relevant to the HB 920 erosion story.

```{r}
#| label: fig-collections-trend
#| fig-cap: "Wisconsin tax collections trend: local property tax, individual income tax, state sales and use tax, and corporate income tax (nominal $)."
trend_long <- readr::read_csv(here("data/intermediate/wi_four_tax_trend.csv"), show_col_types = FALSE)

gfd_trend <- trend_long |>
  select(year, tax_type, collections_usd) |>
  tidyr::pivot_wider(names_from = tax_type, values_from = collections_usd) |>
  mutate(total_taxes = rowSums(across(c(`Local Property Tax`, `Individual Income Tax`, `State Sales and Use Tax`, `Corporate Income Tax`)), na.rm = TRUE))

ymin <- min(trend_long$year, na.rm = TRUE)
ymax <- max(trend_long$year, na.rm = TRUE)

p_trend <- ggplot(trend_long, aes(year, collections_b, color = tax_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  scale_color_manual(values = c(
    "Local Property Tax" = "#1F3864",
    "Individual Income Tax" = "#70AD47",
    "State Sales and Use Tax" = "#5B9BD5",
    "Corporate Income Tax" = "#ED7D31"
  ), name = NULL) +
  scale_x_continuous(breaks = seq(ymin, ymax, 2)) +
  scale_y_continuous(labels = label_dollar(suffix = "B"), expand = expansion(mult = c(0.02, 0.08))) +
  labs(title = "Wisconsin major tax collections over time",
       subtitle = "Local property tax + state individual income, sales/use, and corporate income taxes",
       x = NULL, y = NULL,
       caption = "Source: WIStateandLocalTaxes Tableau workbook (local + state series). Nominal dollars; no inflation adjustment.") +
  theme_swank() +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
p_trend
ggsave(file.path(fig_dir, "fig28_collections_trend_2005_2023.png"),
       p_trend, width = 9, height = 5.5, dpi = 300, bg = "white")

trend_indexed <- trend_long |>
  group_by(tax_type) |>
  mutate(
    base_1985 = collections_b[match(1985, year)],
    index_1985 = 100 * (collections_b / base_1985)
  ) |>
  ungroup()

p_trend_idx1985 <- ggplot(trend_indexed, aes(year, index_1985, color = tax_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  scale_color_manual(values = c(
    "Local Property Tax" = "#1F3864",
    "Individual Income Tax" = "#70AD47",
    "State Sales and Use Tax" = "#5B9BD5",
    "Corporate Income Tax" = "#ED7D31"
  ), name = NULL) +
  scale_x_continuous(breaks = seq(ymin, ymax, 2)) +
  scale_y_continuous(labels = label_number(accuracy = 1), expand = expansion(mult = c(0.02, 0.08))) +
  labs(title = "Wisconsin major tax collections indexed to 1985",
       subtitle = "1985 = 100 for each series (nominal dollars)",
       x = NULL, y = "Index (1985 = 100)",
       caption = "Source: WIStateandLocalTaxes Tableau workbook (local + state series).") +
  theme_swank() +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
p_trend_idx1985
ggsave(file.path(fig_dir, "fig30_collections_trend_indexed_1985.png"),
       p_trend_idx1985, width = 9, height = 5.5, dpi = 300, bg = "white")
```


## The geography of collections

Where the property tax is charged tracks property wealth and population:

```{r}
#| label: fig-county-property
#| fig-cap: "Wisconsin county local property-tax levy, aggregated from county + city + village + town jurisdictions (latest year)."
wi_local_jur <- arrow::read_parquet(
  here("data/raw/02_Local_Government_Dashboard_0.1.5a.twb_Files_federated_1/Extract__Extract.parquet")
) |>
  transmute(
    year = as.integer(format(as.Date(Year), "%Y")),
    county = str_to_title(as.character(`County Name`)),
    muni_type_raw = as.character(`Muni Type Code`),
    local_taxes = suppressWarnings(as.numeric(`Local Taxes (SOT TVC or County)`))
  ) |>
  mutate(
    jurisdiction_type = case_when(
      str_to_upper(muni_type_raw) %in% c("COUNTY") ~ "County",
      str_to_upper(muni_type_raw) %in% c("C", "CITY") ~ "City",
      str_to_upper(muni_type_raw) %in% c("V", "VILLAGE") ~ "Village",
      str_to_upper(muni_type_raw) %in% c("T", "TOWN") ~ "Town",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(year), !is.na(county), !is.na(local_taxes), !is.na(jurisdiction_type))

latest_local_year <- max(wi_local_jur$year, na.rm = TRUE)

wi_county_local_tax <- wi_local_jur |>
  filter(year == latest_local_year) |>
  group_by(county) |>
  summarise(
    property_tax = sum(local_taxes, na.rm = TRUE),
    n_jurisdictions = n(),
    .groups = "drop"
  )

wi_county_map <- wi_county |>
  mutate(county = str_to_title(NAME)) |>
  left_join(wi_county_local_tax, by = "county") |>
  mutate(property_tax = coalesce(property_tax, 0))

p_fig7 <- wi_county_map |>
  ggplot() +
  geom_sf(aes(fill = property_tax), color = "white", linewidth = 0.1) +
  scale_fill_swank_seq(name = "Local property tax levy", labels = label_usd_millions()) +
  labs(
    title = "Where local property-tax levy is concentrated in Wisconsin",
    subtitle = paste0(
      "County totals aggregating county + city + village + town jurisdictions, ",
      latest_local_year
    ),
    caption = "Source: Wisconsin Local Government Dashboard Tableau extract (field: Local Taxes (SOT TVC or County)); summed to county."
  ) +
  theme_swank_map()
p_fig7
ggsave(file.path(fig_dir, "fig07_wi_county_local_property_tax_latest.png"),
       p_fig7, width = 8, height = 7, dpi = 300, bg = "white")
```

The leading replacement candidate — the sales tax — is collected somewhere quite different. This is the **place-of-sale vs place-of-residence** distinction that is central to Layer 2: where a county *collects* sales tax is not where its residents *live* or where the lost property tax was *spent*.

```{r}
#| label: fig-county-sales
#| fig-cap: "Wisconsin county sales-tax distributions (latest full year, place of sale)."
sales_use <- readr::read_csv(
  here("data/intermediate/fig6_wi_county_sales_collections_latest_full_year.csv"),
  show_col_types = FALSE
) |>
  transmute(
    county = str_replace(str_to_title(as.character(county)), "^Saint\\s+", "St. "),
    year = suppressWarnings(as.integer(year)),
    collections = suppressWarnings(as.numeric(collections))
  )

latest_sales_year <- max(sales_use$year, na.rm = TRUE)

wi_county_sales_map <- wi_county |>
  mutate(county = str_to_title(NAME)) |>
  left_join(
    sales_use |>
      filter(year == latest_sales_year) |>
      select(county, collections),
    by = "county"
  ) |>
  mutate(collections = coalesce(collections, 0))

p_fig6 <- wi_county_sales_map |>
  ggplot() +
  geom_sf(aes(fill = collections), color = "white", linewidth = 0.1) +
  scale_fill_swank_seq(name = "Sales-tax distributions", labels = label_usd_millions()) +
  labs(
    title = "Sales-tax collections concentrate where the selling happens",
    subtitle = paste0("Wisconsin county sales-tax distributions, calendar year ", latest_sales_year),
    caption = "Source: Wisconsin DOR County Sales Tax Distributions Tableau workbook. Place-of-sale distributions by county."
  ) +
  theme_swank_map()
p_fig6
ggsave(file.path(fig_dir, "fig06_wi_county_sales_tax_latest_full_year.png"),
       p_fig6, width = 8, height = 7, dpi = 300, bg = "white")
```

Finally, who *bears* the property tax locally depends on the class composition of each county's base — residential/agricultural (Class I) vs commercial, industrial, and mineral (Class II; public-utility tangible property is the separate PUT instrument, not Class II real property).

Based on share of taxable value:

```{r}
#| label: fig-residential-share
#| fig-cap: "Wisconsin county residential share of estimated property-tax collections (latest matched year)."
class_est <- readr::read_csv(
  here("data/intermediate/wi_property_tax_class_estimates_from_millage.csv"),
  show_col_types = FALSE
) |>
  transmute(
    year = suppressWarnings(as.integer(year)),
    county = str_to_title(str_remove(as.character(county_name), "\\s+County$")),
    class_group = as.character(class_group),
    has_rate = as.logical(has_rate),
    est_net_tax = suppressWarnings(as.numeric(estimated_net_tax_from_rate))
  ) |>
  filter(!is.na(year), !is.na(county), !is.na(class_group), !is.na(est_net_tax), has_rate)

latest_fig10_year <- max(class_est$year, na.rm = TRUE)

county_class <- class_est |>
  filter(year == latest_fig10_year) |>
  group_by(county, class_group) |>
  summarise(est_net_tax = sum(est_net_tax, na.rm = TRUE), .groups = "drop")

county_totals <- county_class |>
  group_by(county) |>
  summarise(total_est_net_tax = sum(est_net_tax, na.rm = TRUE), .groups = "drop")

county_residential <- county_class |>
  filter(class_group == "Residential") |>
  select(county, residential_est_net_tax = est_net_tax)

fig10_df <- county_totals |>
  left_join(county_residential, by = "county") |>
  mutate(
    residential_est_net_tax = coalesce(residential_est_net_tax, 0),
    residential_share = if_else(total_est_net_tax > 0, residential_est_net_tax / total_est_net_tax, NA_real_)
  )

p_fig10 <- wi_county |>
  mutate(county = str_to_title(NAME)) |>
  left_join(fig10_df, by = "county") |>
  ggplot() +
  geom_sf(aes(fill = residential_share), color = "white", linewidth = 0.1) +
  scale_fill_gradientn(
    colours = c("#fffde7", "#cfe8ff", "#1f4e8c", "#6a1b9a"),
    values = scales::rescale(c(0, 0.30, 0.70, 1)),
    limits = c(0, 1),
    labels = label_percent(accuracy = 1),
    name = "Residential share",
    na.value = "grey90"
  ) +
  labs(
    title = "How much of each county's estimated property tax comes from residential class",
    subtitle = paste0("Residential share of estimated net property-tax collections, calendar year ", latest_fig10_year),
    caption = "Source: WI AssessedValues Tableau + TVC millage/rate extract. Estimated from assessed value by class × district net millage; pre-credit approximation."
  ) +
  theme_swank_map()
p_fig10
ggsave(file.path(fig_dir, "fig10_residential_share_value_2024.png"),
       p_fig10, width = 8, height = 7, dpi = 300, bg = "white")
```

# (iv) What does each level spend?

Collections are only half the ledger. Using the Government Finance Database (Pierson, Hand & Thompson 2015; Ohio 2022, a Census year), we can see what each local level actually funds — and therefore what loses its revenue source under abolition.

```{r}
#| label: fig-expenditure
#| fig-cap: "What each local government spends on, Ohio 2022 (share of expenditure; total $B labelled)."
exp <- readr::read_csv(here("data/clean/wisconsin_expenditure_by_jurisdiction_2022.csv"),
                       show_col_types = FALSE) |>
  group_by(govt_type) |>
  mutate(total_b = sum(expenditure_usd) / 1e9, share = expenditure_usd / sum(expenditure_usd)) |>
  ungroup()

gt_order  <- exp |> distinct(govt_type, total_b) |> arrange(desc(total_b)) |> pull(govt_type)
cat_order <- exp |> filter(category != "Other") |> group_by(category) |>
  summarise(s = sum(expenditure_usd), .groups = "drop") |> arrange(s) |> pull(category)
exp <- exp |> mutate(govt_type = factor(govt_type, levels = gt_order),
                     category  = factor(category, levels = c("Other", cat_order)))

exp_cols <- c(
  "Education" = "#1F3864", "Public welfare" = "#C55A11", "Health & hospitals" = "#9DC3E6",
  "Highways & roads" = "#A6611A", "Police" = "#BB0000", "Fire protection" = "#E15759",
  "Corrections" = "#843C0C", "Sewerage" = "#76B7B2",
  "Utilities (water/electric/transit)" = "#5B9BD5", "Parks & recreation" = "#70AD47",
  "Housing & community dev" = "#7030A0", "Government administration" = "#2E7D32",
  "Interest on debt" = "#595959", "Other" = "#BFBFBF"
)

p_exp <- ggplot(exp, aes(govt_type, share, fill = category)) +
  geom_col(width = 0.72, color = "white", linewidth = 0.3) +
  geom_text(data = distinct(exp, govt_type, total_b),
            aes(govt_type, 1.01, label = paste0("$", formatC(total_b, 1, format = "f"), "B")),
            inherit.aes = FALSE, vjust = 0, size = 3, fontface = "bold", color = "#404040") +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.08))) +
  scale_fill_manual(values = exp_cols, name = NULL) +
  labs(title = "What each local government spends on, Wisconsin 2022",
       subtitle = "Share of total expenditure by function; total expenditure labelled above each bar",
       x = NULL, y = NULL,
       caption = "Source: Government Finance Database Midwest extract (midwest_2005_present.parquet) via 1c-expenditure-summary.R. All major Census functions broken out; residual in 'Other'.") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank(), legend.text = element_text(size = 8))
p_exp
ggsave(file.path(fig_dir, "fig15_expenditure_by_jurisdiction_2022.png"),
       p_exp, width = 9, height = 6, dpi = 300, bg = "white")


exp <- readr::read_csv(
  here("data/clean/wisconsin_expenditure_by_jurisdiction_2022.csv"),
  show_col_types = FALSE
) |>
  group_by(category) |>
  summarise(
    expenditure_usd = sum(expenditure_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    total_b = sum(expenditure_usd) / 1e9,
    share = expenditure_usd / sum(expenditure_usd)
  )


# Order categories by statewide expenditure,
# while keeping "Other" at the bottom of the stacked bar
cat_order <- exp |>
  filter(category != "Other") |>
  arrange(expenditure_usd) |>
  pull(category)

exp <- exp |>
  mutate(
    category = factor(
      category,
      levels = c("Other", cat_order)
    )
  )


exp_cols <- c(
  "Education" = "#1F3864",
  "Public welfare" = "#C55A11",
  "Health & hospitals" = "#9DC3E6",
  "Highways & roads" = "#A6611A",
  "Police" = "#BB0000",
  "Fire protection" = "#E15759",
  "Corrections" = "#843C0C",
  "Sewerage" = "#76B7B2",
  "Utilities (water/electric/transit)" = "#5B9BD5",
  "Parks & recreation" = "#70AD47",
  "Housing & community dev" = "#7030A0",
  "Government administration" = "#2E7D32",
  "Interest on debt" = "#595959",
  "Other" = "#BFBFBF"
)


p_exp <- ggplot(
  exp,
  aes(
    x = "Ohio",
    y = share,
    fill = category
  )
) +
  geom_col(
    width = 0.55,
    color = "white",
    linewidth = 0.3
  ) +
  geom_text(
    data = distinct(exp, total_b),
    aes(
      x = "Ohio",
      y = 1.01,
      label = paste0(
        "$",
        formatC(total_b, digits = 1, format = "f"),
        "B"
      )
    ),
    inherit.aes = FALSE,
    vjust = 0,
    size = 3,
    fontface = "bold",
    color = "#404040"
  ) +
  scale_y_continuous(
    labels = scales::label_percent(),
    expand = expansion(mult = c(0, 0.08))
  ) +
  scale_fill_manual(
    values = exp_cols,
    name = NULL
  ) +
  labs(
    title = "What Wisconsin local governments spend on, 2022",
    subtitle = "Share of total statewide local government expenditure by function",
    x = NULL,
    y = NULL,
    caption = paste(
      "Source: Government Finance Database Midwest extract",
      "(midwest_2005_present.parquet) via 1c-expenditure-summary.R.",
      "All major Census functions broken out; residual in 'Other'."
    )
  ) +
  theme_swank() +
  theme(
    panel.grid.major.x = element_blank(),
    legend.text = element_text(size = 8)
  )

p_exp
ggsave(file.path(fig_dir, "fig15_expenditure_aggregated_2022.png"),
       p_exp, width = 9, height = 6, dpi = 300, bg = "white")
```

**What it shows.** Schools spend almost entirely on **education**; counties on **public welfare, health, and administration** (Ohio counties run Medicaid and human-services functions); municipalities on **utilities, safety, and roads**; townships on **roads and fire** (fire is its own broken-out function here — townships and municipalities run most of it). The point for the ban: the property tax that funds these is not interchangeable across levels. A township or special district that loses it cannot pivot to a sales or income tax — it has neither (section ii).

**Which services are most at risk?** A first, deliberately simple way to combine the two halves of the ledger: weight each level's spending by how reliant that level is on the property tax (Property_Tax ÷ Total_Revenue, from the GFD), then total by function. It is ad hoc — it assumes the property tax funds every function within a level proportionally — but it gives a transparent first ranking of what is most exposed.

```{r}
#| label: fig-services-at-risk
#| fig-cap: "Estimated property-tax-funded expenditure by function, Ohio 2022 (expenditure × level reliance; ad hoc)."
reliance_gfd <- read_parquet(
  here("data/raw/govt_finance_database/midwest_2005_present.parquet"),
  col_select = all_of(c("state_name", "year", "govt_type_label", "Property_Tax", "Total_Revenue"))
) |>
  filter(state_name == "Wisconsin", year == 2022) |>
  group_by(govt_type_label) |>
  summarise(reliance = sum(Property_Tax, na.rm = TRUE) / sum(Total_Revenue, na.rm = TRUE), .groups = "drop")

at_risk <- readr::read_csv(here("data/clean/wisconsin_expenditure_by_jurisdiction_2022.csv"),
                           show_col_types = FALSE) |>
  left_join(reliance_gfd, by = c("govt_type" = "govt_type_label")) |>
  mutate(at_risk_usd = expenditure_usd * reliance) |>
  group_by(category) |>
  summarise(at_risk_b = sum(at_risk_usd, na.rm = TRUE) / 1e9, .groups = "drop") |>
  arrange(at_risk_b) |>
  mutate(category = factor(category, levels = category))

p_risk <- ggplot(at_risk, aes(at_risk_b, category)) +
  geom_col(fill = swank_palette[1], width = 0.72) +
  geom_text(aes(label = paste0("$", formatC(at_risk_b, 1, format = "f"), "B")),
            hjust = -0.15, size = 3, color = "#404040") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(title = "Education is by far the most property-tax-dependent service",
       subtitle = "Estimated property-tax-funded expenditure by function, Wisconsin 2022",
       x = NULL, y = NULL,
       caption = "Source: GFD reliance (Property_Tax / Total_Revenue by level) × expenditure by function. Ad hoc; assumes proportional funding within a level.") +
  theme_swank() +
  theme(panel.grid.major.y = element_blank(), axis.text.x = element_blank(),
        panel.grid.major.x = element_blank())
p_risk
ggsave(file.path(fig_dir, "fig19_services_at_risk_2022.png"),
       p_risk, width = 8, height = 5, dpi = 300, bg = "white")
```

By this rough cut, **education (\~\$12B)** dwarfs everything — schools spend almost entirely on instruction and are \~42% property-tax-reliant — followed distantly by **government administration (\~\$1.4B)**, then a cluster of **police and public welfare (\~\$0.6B each)** and **fire protection and roads (\~\$0.5B each)**, with **sewerage (\~\$0.3B)** close behind. (Fire and sewerage are now broken out from "Other," which is why they surface here.) The reliance weights also surface a vulnerability the collections totals hide: **townships are the most property-tax-reliant level (≈60% of their revenue) and school districts ≈42%**, while municipalities (≈7%) could lean on their income tax instead. So the services most exposed are schools first, then the township basics — roads and fire — that have no fallback tax. *Treat the dollar levels as indicative pending the ODT / ODE&W reconciliation.*

# Bringing in the ACS data

The tables above are *fiscal* (dollars by jurisdiction). Turning them into *who-bears-it* requires the ACS household data (`data/clean/acs_county_wi_2024_5yr.parquet` and `pums_households_oh.parquet`, script `1a`). Two joins do most of the work:

- **County denominators.** Dividing the county property-tax-charged map (§iii) by ACS households or population gives property tax *per household* and *per capita* — the denominators every redistribution formula needs (Layer 2), and a fairer way to compare counties than raw dollars.


The county-denominator join comes first. Dividing county property tax charged (§iii) by ACS households reframes the map from "where the dollars are" to "what a typical household is charged" — a fairer basis for comparison and the denominator Layer 2 needs:

```{r}
#| label: fig-per-household
#| fig-cap: "County property tax charged per household (TY2025 ÷ ACS households)."
acs_cty <- read_parquet(here("data/clean/acs_county_wi_2024_5yr.parquet")) |>
  select(county, households, population)

# per_hh <- prop_county |>
#   left_join(acs_cty, by = "county") |>
#   mutate(prop_per_hh = prop_tax / households)

# p_perhh <- oh_counties |> left_join(per_hh, by = "county") |>
#   ggplot() +
#   geom_sf(aes(fill = prop_per_hh), color = "white", linewidth = 0.1) +
#   scale_fill_swank_seq(name = "Property tax / household", labels = label_dollar()) +
#   labs(title = "Property tax charged per household",
#        subtitle = "County property tax charged (TY2025) ÷ ACS households (2020-2024 5-year)",
#        caption = "Source: ODT Table PD-23 (pd23ty25.xlsx) ÷ ACS 5-yr households (acs_county_wi_2024_5yr.parquet).") +
#   theme_swank_map()
# p_perhh
# ggsave(file.path(fig_dir, "fig16_property_tax_per_household.png"),
#        p_perhh, width = 8, height = 7, dpi = 300, bg = "white")
```

This per-household view ranks counties very differently from the raw-dollar map: wealthy suburban counties rise, and the metro counties fall once you divide by their many households — a preview of why the redistribution layer's *denominator choice* (per household vs per capita vs per pupil) materially changes who looks like a winner. Two more denominators make the point directly:

```{r}
# #| label: fig-per-capita
# #| fig-cap: "County property tax charged per capita (TY2025 ÷ ACS population)."
# per_cap <- prop_county |>
#   left_join(select(acs_cty, county, population), by = "county") |>
#   mutate(prop_per_capita = prop_tax / population)

# p_percap <- oh_counties |> left_join(per_cap, by = "county") |>
#   ggplot() +
#   geom_sf(aes(fill = prop_per_capita), color = "white", linewidth = 0.1) +
#   scale_fill_swank_seq(name = "Property tax / capita", labels = label_dollar()) +
#   labs(title = "Property tax charged per capita",
#        subtitle = "County property tax charged (TY2025) ÷ ACS population (2020-2024 5-year)",
#        caption = "Source: ODT Table PD-23 (pd23ty25.xlsx) ÷ ACS 5-yr population (acs_county_wi_2024_5yr.parquet).") +
#   theme_swank_map()
# p_percap
# ggsave(file.path(fig_dir, "fig30_property_tax_per_capita.png"),
#        p_percap, width = 8, height = 7, dpi = 300, bg = "white")
```

Per-capita spreads counties out differently again from per-household (household size varies by county, so the two denominators don't just rescale each other) — smaller-household, older counties rank higher per capita than they do per household.

The third denominator, **per pupil**, only makes sense at the *school-district* level (property tax funds schools, not counties, and Ohio's \~610 districts don't nest inside counties). The GFD carries district-level enrollment directly, so no separate NCES/ODE&W pull is needed for a first cut — but there is no crosswalk yet from GFD's own government identifiers to the validated ODT-IRN/TIGER-GEOID crosswalk used elsewhere in this document (`xwalk`), so this is a distribution, not a map, until that crosswalk exists:

```{r}
#| label: fig-per-pupil
#| fig-cap: "Distribution of school-district property tax per pupil, Ohio 2022 (GFD)."
gfd_sd <- read_parquet(
  here("data/raw/govt_finance_database/midwest_2005_present.parquet"),
  col_select = all_of(c("state_name", "year", "govt_type_label", "Enrollment", "Property_Tax"))
) |>
  filter(state_name == "Wisconsin", year == 2022, govt_type_label == "School District", Enrollment > 0) |>
  mutate(property_tax = Property_Tax * 1000, per_pupil = property_tax / Enrollment)

statewide_per_pupil <- sum(gfd_sd$property_tax) / sum(gfd_sd$Enrollment)

p_perpupil <- ggplot(gfd_sd, aes(per_pupil)) +
  geom_histogram(bins = 40, fill = swank_palette[1], color = "white", linewidth = 0.2) +
  geom_vline(xintercept = statewide_per_pupil, linetype = "dashed", color = swank_palette[2], linewidth = 0.6) +
  annotate("text", x = statewide_per_pupil, y = Inf, label = paste0("Statewide: ", label_dollar()(round(statewide_per_pupil))),
          hjust = -0.05, vjust = 1.4, size = 3.2, color = swank_palette[2]) +
  scale_x_continuous(labels = label_dollar(), limits = c(0, quantile(gfd_sd$per_pupil, 0.99))) +
  labs(title = "Property tax per pupil varies enormously across Ohio school districts",
       subtitle = paste0(nrow(gfd_sd), " districts with reported enrollment, Ohio 2022 (top 1% trimmed from the x-axis for readability)"),
       x = "Property tax per pupil", y = "Number of districts",
       caption = "Source: Government Finance Database, Wisconsin school districts, 2022. Districts with a large commercial/industrial/utility base relative to enrollment (e.g. a power plant) sit far out on the right tail -- real dispersion, not a data error.") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank())
p_perpupil
ggsave(file.path(fig_dir, "fig31_property_tax_per_pupil_distribution.png"),
       p_perpupil, width = 8, height = 5, dpi = 300, bg = "white")
```

The statewide average is about **`r label_dollar(accuracy = 100)(statewide_per_pupil)`** per pupil, but individual districts range from near \$0 to well over ten times that — the property-wealthy-vs-property-poor divide that Ohio's school-funding litigation (*DeRolph*) and equalization formulas exist to address, and exactly the kind of variation the Layer-3 school-finance work needs to carry forward, not average away.

A useful consistency check connects the two joins: the owner-occupied residential property tax implied by PUMS should reconcile against the residential share of the PD-23 charged totals, and (eventually) against the CoreLogic parcel data. The ACS-vs-ODT comparison figure further below performs exactly that check.

# Household incidence: who currently bears the property tax (Layer 4 preview)

The maps so far are about *places*. The PUMS file lets us ask the distributional question directly: as a share of income, who pays the property tax now? Among owner-occupants the answer is the textbook one — the property tax takes a larger bite of lower incomes.

```{r}
#| label: fig-incidence-owner
#| fig-cap: "Owner-occupant property tax as a share of household income, by income decile."
pums  <- read_parquet(here("data/clean/pums_households_wi.parquet"))
inc_q <- property_tax_incidence(pums, n_groups = 10L)

p_inc <- inc_q |>
  mutate(decile = factor(grp, levels = 1:10,
           labels = c("D1\n(lowest)", "D2", "D3", "D4", "D5", "D6", "D7", "D8", "D9", "D10\n(highest)"))) |>
  ggplot(aes(decile, eff_rate_agg)) +
  geom_col(fill = swank_palette[1], width = 0.7) +
  geom_text(aes(label = percent(eff_rate_agg, accuracy = 0.1)),
            vjust = -0.4, size = 3.6, color = "#404040") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Property tax takes a larger share of income for lower-income owners",
       subtitle = "Owner-occupant property tax as a share of household income, by income decile (constant 2024$)",
       x = "Household income decile (weighted)", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS (pums_households_oh.parquet), owner-occupants. Aggregate tax/income; annual basis.") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank())
p_inc
ggsave(file.path(fig_dir, "fig17_pums_property_incidence_decile.png"),
       p_inc, width = 8, height = 5, dpi = 300, bg = "white")
```

Three caveats travel with this figure, and all three matter for the model:

- **Owners only, so far.** Renters do not report a property-tax bill in PUMS, but they bear part of it through rents (pass-through). We add a first-cut renter estimate immediately below and combine the two.
- **Annual income, not lifetime.** The lowest quintile is inflated by asset-rich, low-current-income owners (retirees, a bad-income year). That is why we report the *aggregate* rate (weighted total tax ÷ weighted total income), not the mean of household ratios — which exceeds 100% at the very bottom. Read Q1 as "elevated," not as a precise number.
- **This is the *current* burden.** Its only job in the model is to be the baseline the replacement scenarios are measured against: a sales-tax replacement would steepen the bottom (consumption is a larger income share for low earners), an income-tax replacement would flatten or reverse it. That comparison is Layer 4 proper; this figure is its foundation.

## Adding renters, and the combined picture

Renters bear property tax indirectly — landlords pass part of it through into rent. To get a renter burden we impute the property tax on each rental unit from its rent and assign the tenant a pass-through share. This figure fixes two inputs for that imputation: a 50% pass-through share and a price-to-rent ratio of 16, chosen when this baseline figure was first built.

```{r}
#| label: fig-incidence-combined
#| fig-cap: "Property tax as a share of income by quantile — owners, renters (50% pass-through), and combined."
pums_b <- add_ptax_burden(pums, p2r = 16, pass_through = 0.50)
inc_all <- bind_rows(
  incidence_by_who(pums_b, "owner",    5L),
  incidence_by_who(pums_b, "renter",   5L),
  incidence_by_who(pums_b, "combined", 5L)
)

# Whole-population (not by-quantile) aggregate rate for owners vs. renters --
# n_groups = 1L collapses wtd_quantile_bins() to a single group, so
# eff_rate_agg becomes the population-wide weighted-mean-tax / weighted-
# mean-income ratio for that tenure. Computed live so the prose below never
# goes stale under a data refresh (e.g. an ACS vintage change).
owner_agg_rate  <- incidence_by_who(pums_b, "owner",  1L)$eff_rate_agg
renter_agg_rate <- incidence_by_who(pums_b, "renter", 1L)$eff_rate_agg

p_inc3 <- inc_all |>
  mutate(quantile = factor(grp, levels = 1:5, labels = paste0("Q", 1:5)),
         who = factor(who, levels = c("owner", "renter", "combined"),
                      labels = c("Owners", "Renters (50% pass-through)", "Combined"))) |>
  ggplot(aes(quantile, eff_rate_agg, fill = who)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.74) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.08))) +
  scale_fill_manual(values = c("Owners" = swank_palette[1],
                               "Renters (50% pass-through)" = swank_palette[4],
                               "Combined" = swank_palette[3]), name = NULL) +
  labs(title = "Property tax is regressive on income for owners and renters alike",
       subtitle = "Property tax as a share of household income by income quantile (constant 2024$)",
       x = "Household income quantile (weighted)", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS. Renter burden imputed (50% pass-through, price-to-rent 16) — provisional.") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
p_inc3
ggsave(file.path(fig_dir, "fig20_pums_incidence_owner_renter_combined.png"),
       p_inc3, width = 9, height = 5, dpi = 300, bg = "white")
```

Owners and renters both show the regressive-on-income shape. Renters' burden is a touch lower in aggregate (`r percent(renter_agg_rate, accuracy = 0.1)` vs `r percent(owner_agg_rate, accuracy = 0.1)`) — partly real, partly the 50% assumption — but **at the bottom of the distribution renters are hit comparably hard**, because rent is a large share of a low income. The combined bars are the closest thing here to a whole-population view of who bears Ohio's property tax today, and they are the baseline the replacement scenarios get judged against in Layer 4.

## The other half of the story: total dollars paid

The share-of-income view says the property tax is regressive. The *dollar* view says the opposite-sounding-but-also-true thing: the top of the distribution pays the most in total, because higher-income households occupy more valuable housing. Both matter — the rate drives the equity story, the dollar level drives the revenue / who-funds-what story. Here it is by income **quintile**:

```{r}
#| label: fig-incidence-total-dollars
#| fig-cap: "Estimated total property tax paid by income quantile — owners, renters (50% pass-through), combined."
paid_all <- bind_rows(
  incidence_by_who(pums_b, "owner",    5L),
  incidence_by_who(pums_b, "renter",   5L),
  incidence_by_who(pums_b, "combined", 5L)
)

p_paid <- paid_all |>
  mutate(quantile = factor(grp, levels = 1:5, labels = paste0("Q", 1:5)),
         who = factor(who, levels = c("owner", "renter", "combined"),
                      labels = c("Owners", "Renters (50% pass-through)", "Combined")),
         total_b = total_paid / 1e9) |>
  ggplot(aes(quantile, total_b, fill = who)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.74) +
  geom_text(aes(label = paste0("$", formatC(total_b, 1, format = "f"), "B")),
            position = position_dodge(width = 0.8), vjust = -0.3, size = 2.7, color = "#404040") +
  scale_y_continuous(labels = label_dollar(suffix = "B"), expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Owners" = swank_palette[1],
                               "Renters (50% pass-through)" = swank_palette[4],
                               "Combined" = swank_palette[3]), name = NULL) +
  labs(title = "In total dollars, the highest-income quantile pays the most property tax",
       subtitle = "Estimated total annual property tax paid by income quantile (constant 2024$)",
       x = "Household income quantile (weighted)", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS. TAXAMT top-coded at $27,500 -> top quintile understated. Renter burden imputed (provisional).") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
p_paid
ggsave(file.path(fig_dir, "fig21_pums_total_paid_quantile.png"),
       p_paid, width = 9, height = 5, dpi = 300, bg = "white")
```

The figure below checks that same PUMS-based aggregate directly against the ODT administrative totals — a magnitude reconciliation, not an incidence comparison.

```{r}
#| label: fig-acs-vs-odt-total
#| fig-cap: "Total residential property tax paid: ACS PUMS combined (owner+renter) imputed estimate vs. ODT RPU-Abs residential-class levied and net totals, TY2023."
paid_all_10 <- bind_rows(
  incidence_by_who(pums_b, "owner",    10L),
  incidence_by_who(pums_b, "renter",   10L),
  incidence_by_who(pums_b, "combined", 10L)
)

# Only the "combined" rows are summed here. incidence_by_who(..., "combined")
# already pools every owner AND renter household into one set of decile
# rows, so summing total_paid across all three `who` levels double-counted
# owner+renter dollars a second time via "combined" (~$27B instead of the
# correct ~$13.5B). Filtering to "combined" alone is the right ACS-PUMS
# aggregate -- see 05_output/tables/pums_property_total_paid_decile.csv for
# the reconciled owner/renter/combined totals this mirrors.
total_paid <- paid_all_10 |>
  filter(who == "combined") |>
  summarise(total_paid_resi = sum(total_paid, na.rm = TRUE)/1000000000)

# Decile-grain owner totals, used in the "total dollars" paragraph below --
# the D10-specific figure needs decile (10L) grain, which only exists here;
# fig21 above (the quantile-grain "total dollars" figure) is Q1-Q5.
owner_d10_b   <- sum(paid_all_10$total_paid[paid_all_10$who == "owner" & paid_all_10$grp == 10], na.rm = TRUE) / 1e9
owner_total_b <- sum(paid_all_10$total_paid[paid_all_10$who == "owner"], na.rm = TRUE) / 1e9

property_tax_total <- read_excel(
  odt("Local Tax/Property_Tax_Real_Property/rpuabs23.xlsx"),
  skip = 3,
  sheet = "RPUAbs23"
) |>
  filter(`Field Name` == "Current real and public utility taxes levied") %>% summarise(total_paid = sum(as.numeric(`Total All Property`), na.rm = TRUE)/1000000000,
  total_paid_resi = sum(as.numeric(`Residential`), na.rm = TRUE)/1000000000)

property_tax_total_minus <- read_excel(
  odt("Local Tax/Property_Tax_Real_Property/rpuabs23.xlsx"),
  skip = 3,
  sheet = "RPUAbs23"
) |>
  filter(`Field Name` == "Credit from reduction factors") %>% summarise(total_paid = sum(as.numeric(`Total All Property`), na.rm = TRUE)/1000000000,
  total_paid_resi = sum(as.numeric(`Residential`), na.rm = TRUE)/1000000000)

property_tax_net <- read_excel(
  odt("Local Tax/Property_Tax_Real_Property/rpuabs23.xlsx"),
  skip = 3,
  sheet = "RPUAbs23"
) |>
  filter(`Field Name` == "Net total") %>% summarise(total_paid = sum(as.numeric(`Total All Property`), na.rm = TRUE)/1000000000,
  total_paid_resi = sum(as.numeric(`Residential`), na.rm = TRUE)/1000000000)

combine_all <- bind_rows(
  total_paid %>% mutate(source = "ACS PUMS"),
  property_tax_total %>% mutate(source = "Ohio Tax Data (Levied)"),
  property_tax_net %>% mutate(source = "Ohio Tax Data (Net)")
)

acs_total_resi_b  <- total_paid$total_paid_resi
odt_levied_resi_b <- property_tax_total$total_paid_resi
odt_net_resi_b    <- property_tax_net$total_paid_resi

p_compare_src <- combine_all |>
  ggplot(aes(source, total_paid_resi, fill = source)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.74) +
  geom_text(aes(label = paste0("$", formatC(total_paid_resi, 1, format = "f"), "B")),
            position = position_dodge(width = 0.8), vjust = -0.3, size = 2.7, color = "#404040") +
  scale_y_continuous(labels = label_dollar(suffix = "B"), expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("ACS PUMS" = swank_palette[1],
                               "Ohio Tax Data (Levied)" = swank_palette[4],
                               "Ohio Tax Data (Net)" = swank_palette[3]), name = NULL) +
  labs(title = "ACS PUMS's imputed total tracks ODT's net (post-credit) total, not the pre-credit levy",
       subtitle = "Total annual residential property tax paid, by source (constant 2024$)",
       x = "Source", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS (combined owner+renter, 50% pass-through) vs. ODT RPU-Abs (rpuabs23.xlsx), Residential field, TY2023.") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
p_compare_src
ggsave(file.path(fig_dir, "fig_comparison_source_property_tax_magnitude.png"),
       p_compare_src, width = 9, height = 5, dpi = 300, bg = "white")
```

The corrected ACS PUMS aggregate is `r dollar(acs_total_resi_b, accuracy = 0.1, suffix = "B")`, below ODT's post-credit net total of `r dollar(odt_net_resi_b, accuracy = 0.1, suffix = "B")` and well below the pre-credit levied total of `r dollar(odt_levied_resi_b, accuracy = 0.1, suffix = "B")` — the PUMS-based estimate tracks the *net* (after HB 920 reduction factors and rollbacks) concept, which is the right comparison, since PUMS respondents report what they actually paid. The gap to the net total is broadly consistent with PUMS's own `TAXAMT` top-coding (noted below) and a conservative renter pass-through imputation, both of which understate the true aggregate rather than overstate it. One vintage caveat on this comparison: the "(constant 2024$)" subtitle above describes only the ACS PUMS bar. The ODT RPU-Abs comparator (`rpuabs23.xlsx`, Residential field) is still TY2023 nominal — RPU-Abs is out of scope for the 2026-07-20 ACS-vintage migration (it stays on TY2023 for the separate, PI-locked renter relief-share gross-up elsewhere in this document) — so this figure now compares a 2024$ series against an unchanged TY2023 series, a roughly one-year-wider vintage gap than before the migration (PD-23's own audited Class I growth rate is \~4.66%/year, `05_output/audit/sensitivity_ptax_vintage_audit.csv`). That widening works in the same direction as the two understatement sources above, so it does not change the qualitative reading, but it is part of why the gap to the net total looks as large as it does.

So the two incidence views are complements. As a *share of income* the property tax falls hardest on lower-income households; in *total dollars* the top decile pays the most (owners' D10 alone ≈ `r dollar(owner_d10_b, accuracy = 0.01, suffix = "B")` of the `r dollar(owner_total_b, accuracy = 0.01, suffix = "B")` owner total, decile grain from the ACS-vs-ODT comparison above). The model will report both. One caveat to carry: PUMS top-codes property tax at \$27,500, so the top decile's true total is somewhat higher than shown — the dollar gradient is if anything steeper.

## A first move toward income-tax incidence

The PUMS pull's person-level income lets us ask the flip side of the property-tax question: how would an **income tax** distribute? This subsection and the two that follow build that comparison in three steps of increasing rigor. This subsection applies Ohio's (now near-flat) income tax to PUMS households directly, treating each household as a single filer — a first, deliberately simple cut. The next subsection replaces the household with the real tax-filing unit. The subsection after that replaces the simple flat-rate calculation with `02_code/2a-income-tax-baseline.R`'s full hand-coded TY2026 statutory construction, the baseline Part II's income-replacement scenarios actually use.

```{r}
#| label: fig-property-vs-income
#| fig-cap: "Property tax vs. income tax as a share of household income, by income decile."
persons   <- read_parquet(here("data/clean/pums_persons_wi.parquet"))
tax_units <- ohio_income_tax_units(persons)        # household = tax unit, TY2026 flat law
income_tax_units_total_b <- sum(tax_units$income_tax * tax_units$wgtp, na.rm = TRUE) / 1e9

# Compare the property tax (combined owner+renter, from pums_b above) with the
# income tax, as a share of income, on the same households and income deciles.
hh_both <- pums_b |>
  select(serialno, wgtp, hh_income, ptax_burden) |>
  left_join(select(tax_units, serialno, income_tax), by = "serialno") |>
  filter(hh_income > 0) |>
  mutate(grp = wtd_quantile_bins(hh_income, wgtp, 10L))

compare <- hh_both |>
  group_by(grp) |>
  summarise(
    `Property tax (current)`   = weighted.mean(ptax_burden, wgtp, na.rm = TRUE) / weighted.mean(hh_income, wgtp),
    `Income tax (current law)` = weighted.mean(income_tax,  wgtp, na.rm = TRUE) / weighted.mean(hh_income, wgtp),
    .groups = "drop") |>
  tidyr::pivot_longer(-grp, names_to = "tax", values_to = "rate")
readr::write_csv(compare, here("05_output/tables/pums_property_vs_income_incidence.csv"))

p_pvi <- compare |>
  mutate(decile = factor(grp, levels = 1:10, labels = paste0("D", 1:10))) |>
  ggplot(aes(decile, rate, fill = tax)) +
  geom_col(position = position_dodge(0.8), width = 0.74) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)),
            position = position_dodge(0.8), vjust = -0.3, size = 2.6, color = "#404040") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Property tax (current)" = swank_palette[1],
                               "Income tax (current law)" = swank_palette[5]), name = NULL) +
  labs(title = "Property tax is regressive; the income tax is progressive",
       subtitle = "Tax as a share of household income, by income decile (constant 2024$)",
       x = "Household income decile (weighted)", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS. Property = combined owner+renter (50% pass-through); income = hand-coded Ohio flat 2.75% (crude). Current levels, not revenue-matched.") +
  theme_swank() +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
p_pvi
ggsave(file.path(fig_dir, "fig22_property_vs_income_incidence.png"),
       p_pvi, width = 9, height = 5, dpi = 300, bg = "white")
```

This is the single most important picture for the "who pays instead" question. The **property tax falls hardest on lower-income households** (it tracks housing, a big share of a low income); the **income tax does the opposite**, rising with income. Replacing the property tax with an income tax would therefore shift the burden up the income distribution, toward higher earners — the finding the rest of this income-tax arc exists to sharpen, not overturn.

As a sanity check on the PUMS-to-tax-unit approach, this simple hand-code's weighted statewide total is `r dollar(income_tax_units_total_b, accuracy = 0.1, suffix = "B")`, close to Ohio's actual `r dollar(collections$collections_b[collections$tax_type == "Individual income tax"], accuracy = 0.1, suffix = "B")` individual income tax (TY2024, ODT Table Y-1 — the same figure the collections chart above uses).

Three things to hold, resolved to varying degrees by the two subsections that follow. First, levels are not matched: today's actual income tax raises `r dollar(collections$collections_b[collections$tax_type == "Individual income tax"], accuracy = 0.1, suffix = "B")` against the real property tax's `r dollar(collections$collections_b[collections$tax_type == "Property tax (real, net charged)"], accuracy = 0.1, suffix = "B")`, so this figure compares *shapes*, not dollars — a revenue-neutral replacement would scale the income-tax bars up roughly `r round(collections$collections_b[collections$tax_type == "Property tax (real, net charged)"] / collections$collections_b[collections$tax_type == "Individual income tax"], 1)`-fold while keeping the progressive tilt; the scenario matrix (`scenarios.yml`, Part II) is where we solve for the rate that actually replaces the property tax and report the matched-revenue distribution. Second, this hand-code is crude — household treated as a single tax unit, Ohio taxable income proxied as income minus Social Security/SSI/assistance, the flat statutory rate applied with no personal exemption, no deductions or credits, no federal interaction — and the next two subsections refine it, first by building real tax-filing units, then by substituting 2a's full statutory construction. Third, this is the *current* income tax, not a designed replacement; the scenario matrix reports the *change* in burden a revenue-matched income-tax scenario would produce, not the level shown here.

## Getting more sophisticated: real tax units

The household = tax unit shortcut above is convenient but blunt — a household with two unrelated working roommates is treated as one filer with one exemption, which overstates its tax. The fix is to split each household into actual **filing units** using the relationship and marital-status variables we now pull: the householder and a married spouse file jointly; every other adult files on their own; minors attach as dependents.

```{r}
#| label: fig-income-tax-units
#| fig-cap: "Hand-coded Ohio income tax with real tax units — effective rate by tax-unit income decile."
refined <- ohio_income_tax_taxunits(persons)     # persons read in the section above

ri <- refined |>
  filter(tu_income > 0) |>
  mutate(grp = wtd_quantile_bins(tu_income, wgtp, 10L)) |>
  group_by(grp) |>
  summarise(rate    = weighted.mean(income_tax, wgtp) / weighted.mean(tu_income, wgtp),
            total_b = sum(income_tax * wgtp) / 1e9, .groups = "drop")

# Live reads for the paragraph below, in place of the hand-typed figures it
# used to carry. n_units/n_hh/multi_filer_share are unweighted PUMS-record
# counts (the sample construction, not a population estimate); total_taxunits_b
# and top_rate are weighted, matching every other total in this document.
n_units             <- nrow(refined)
n_hh                <- dplyr::n_distinct(refined$serialno)
multi_filer_share   <- (refined |> dplyr::count(serialno) |> dplyr::filter(n > 1) |> nrow()) / n_hh
total_taxunits_b     <- sum(refined$income_tax * refined$wgtp, na.rm = TRUE) / 1e9
top_rate            <- ri$rate[ri$grp == 10]

p_tu <- ri |>
  mutate(decile = factor(grp, levels = 1:10, labels = paste0("D", 1:10))) |>
  ggplot(aes(decile, rate)) +
  geom_col(fill = swank_palette[5], width = 0.7) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)), vjust = -0.4, size = 3.4, color = "#404040") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
  labs(title = "The income tax stays progressive once we build real tax units",
       subtitle = "Hand-coded Ohio flat tax, by tax-unit income decile (constant 2024$)",
       x = "Tax-unit income decile (weighted)", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS. Tax units = householder+spouse joint, other adults single, minors as dependents. Crude flat 2.75%.") +
  theme_swank() + theme(panel.grid.major.x = element_blank())
p_tu
ggsave(file.path(fig_dir, "fig23_income_tax_units_incidence.png"),
       p_tu, width = 8, height = 5, dpi = 300, bg = "white")
```

Splitting households this way produces **\~`r scales::comma(n_units, accuracy = 1000)` tax units from \~`r scales::comma(n_hh, accuracy = 1000)` households** — about **`r scales::percent(multi_filer_share, accuracy = 1)` of households contain more than one filer** — and because each unit gets its own exemption, the hand-coded total falls from **`r dollar(income_tax_units_total_b, accuracy = 0.1, suffix = "B")`** (household = unit) to about **`r dollar(total_taxunits_b, accuracy = 0.1, suffix = "B")`**. The *shape* is unchanged — still progressive, rising from \~0% at the bottom to **`r scales::percent(top_rate, accuracy = 0.1)`** at the top — but the refinement matters for the *level* (more exemptions claimed) and for *low-income units* (an adult child or roommate now shows up as their own low-income filer rather than being averaged into a higher-income household). This tax-unit construction is the methodologically right base for the flat-tax counterfactual, and it carries forward into the next subsection, which substitutes 2a's full statutory construction for the flat-rate shortcut used here and above.

## The anchored baseline: hand-coded TY2026 law

The simple hand-code above is transparent but misses real tax law — a personal exemption and the statutory base amount. The natural next step would be to run the same tax units through NBER TAXSIM under Ohio's current law, but **TAXSIM cannot do this**: the installed `usincometaxes` package (verified 2026-07-15) errors on any law year after 2023, and no newer CRAN version exists. `02_code/2a-income-tax-baseline.R` instead hand-codes Ohio's full TY2026 statute directly — a single flat 2.75% bracket (\$0 at or below \$26,050 of taxable income, then \$332 plus 2.75% of the excess) with MAGI-tiered personal exemptions (\$2,350 / \$2,100 / \$1,850 per exemption, eliminated entirely at \$500,000 MAGI) — and this is the **anchored income-tax baseline** for the rest of the analysis, including every scenario in Part II. The schedule is validated three ways: the \$332 base amount against two independent sources (the ORC 5747.02 text and a cross-check against a national tax-policy summary), a bracket-continuity diagnostic (the schedule must evaluate identically at each bracket boundary) that pinned the correct \$360.69/\$332 values after a PI review caught a stale-worksheet \$342 base during construction (logged in `MEMORY.md`'s `[LEARN:ohio-statute]` entry), and `02_code/functions/test_ohio_income_handcode.R`'s 19 automated assertions (`make test`). TAXSIM survives in exactly one place in this project — `02_code/3b-income-incidence.R`, for *federal* marginal rates only, as a documented 2023-law stable-structure proxy — and does not touch the Ohio-side liability anywhere.

The figure below therefore compares two hand-coded constructions of Ohio law, not two law years: this document's own simple hand-code (flat 2.75% above \$26,050, no personal exemption, immediately above) against 2a's full statutory construction (personal exemptions plus the \$332 base). Framed this way, the figure is a construction-refinement check on the anchored baseline, not a validation against an independent engine.

```{r}
#| label: fig-taxsim-vs-handcoded
#| fig-cap: "Ohio income tax as a share of income, by tax-unit decile — simple hand-code vs. 2a's full hand-coded TY2026 baseline."
# income_tax_taxsim.csv is written by 02_code/2a-income-tax-baseline.R. The
# filename is legacy (2a called TAXSIM through 2026-07-14); since the
# 2026-07-15 income-handcode phase, 2a no longer calls TAXSIM at all -- its
# `ohio_income_tax` column is 2a's own HAND-CODED TY2026 Ohio liability
# (full bracket schedule + MAGI-tiered personal exemptions). TAXSIM
# survives only in 02_code/3b, for FEDERAL marginal rates. Renamed from
# `taxsim` to `full_2a` here so the object name does not misstate what it
# holds.
full_2a   <- readr::read_csv(here("data/clean/income_tax_taxsim.csv"), show_col_types = FALSE)
handcoded <- ohio_income_tax_taxunits(persons) |> select(tu, hc_tax = income_tax)

# Rank and take rate denominators on tu_income_dist — full per-unit PUMS income
# INCLUDING Social Security (the complete, doc-consistent income concept). Using
# the TAXSIM-input income here would push SS-only retirees into a spurious near-zero
# bottom bin and distort the D1 rate (Codex QC #4).
joined <- full_2a |>
  left_join(handcoded, by = "tu") |>
  filter(tu_income_dist > 0) |>
  mutate(grp = wtd_quantile_bins(tu_income_dist, wgtp, 10L))

cmp <- joined |>
  group_by(grp) |>
  summarise(
    `Simple hand-code (no personal exemption)` = weighted.mean(hc_tax, wgtp, na.rm = TRUE) / weighted.mean(tu_income_dist, wgtp),
    `2a full baseline (TY2026 statute)`        = weighted.mean(ohio_income_tax, wgtp, na.rm = TRUE) / weighted.mean(tu_income_dist, wgtp),
    .groups = "drop") |>
  tidyr::pivot_longer(-grp, names_to = "method", values_to = "rate")
readr::write_csv(cmp, here("05_output/tables/income_tax_handcoded_vs_taxsim.csv"))

# Weighted statewide totals for the two constructions -- read live in the
# prose below rather than hand-typed (additive to this chunk's existing
# computation; does not change what is plotted).
total_simple_b        <- sum(joined$hc_tax * joined$wgtp, na.rm = TRUE) / 1e9
total_2a_b             <- sum(joined$ohio_income_tax * joined$wgtp, na.rm = TRUE) / 1e9
pct_diff_2a_vs_simple  <- (total_2a_b - total_simple_b) / total_simple_b
d10_share_2a           <- sum(joined$ohio_income_tax[joined$grp == 10] * joined$wgtp[joined$grp == 10], na.rm = TRUE) / sum(joined$ohio_income_tax * joined$wgtp, na.rm = TRUE)
d10_share_simple        <- sum(joined$hc_tax[joined$grp == 10] * joined$wgtp[joined$grp == 10], na.rm = TRUE) / sum(joined$hc_tax * joined$wgtp, na.rm = TRUE)

p_ts <- cmp |>
  mutate(decile = factor(grp, levels = 1:10, labels = paste0("D", 1:10))) |>
  ggplot(aes(decile, rate, fill = method)) +
  geom_col(position = position_dodge(0.8), width = 0.74) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)),
            position = position_dodge(0.8), vjust = -0.3, size = 2.6, color = "#404040") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Simple hand-code (no personal exemption)" = "#BFBFBF",
                               "2a full baseline (TY2026 statute)" = swank_palette[5]), name = NULL) +
  labs(title = "The full statutory construction changes the level, not the shape",
       subtitle = "Ohio income tax as a share of income, by tax-unit income decile (constant 2024$)",
       x = "Tax-unit income decile (weighted)", y = NULL,
       caption = "Source: ACS 2020-2024 PUMS. Both series are hand-coded TY2026 Ohio law (02_code/2a no longer calls TAXSIM); the difference is construction, not law year -- simple flat rate vs. 2a's full bracket + exemption schedule.") +
  theme_swank() +
  theme(legend.position = "top", panel.grid.major.x = element_blank())
p_ts
ggsave(file.path(fig_dir, "fig24_taxsim_vs_handcoded.png"),
       p_ts, width = 9, height = 5, dpi = 300, bg = "white")
```

Three results worth keeping. First, the full construction raises the total: 2a's weighted statewide baseline is `r dollar(total_2a_b, accuracy = 0.1, suffix = "B")`, `r percent(abs(pct_diff_2a_vs_simple), accuracy = 1)` `r if (pct_diff_2a_vs_simple > 0) "higher" else "lower"` than the simple hand-code's `r dollar(total_simple_b, accuracy = 0.1, suffix = "B")` (fig23) — the \$332 flat base, owed by every unit above the zero-bracket, outweighs what the personal exemption saves them, and the result sits closer to the `r dollar(collections$collections_b[collections$tax_type == "Individual income tax"], accuracy = 0.1, suffix = "B")` TY2024 administrative total (ODT Table Y-1) than the simple construction does. Second, the shape is progressive under either construction, but concentration at the top is somewhat lower under the full construction: the top decile holds `r percent(d10_share_2a, accuracy = 1)` of 2a's total against `r percent(d10_share_simple, accuracy = 1)` under the simple hand-code, because the flat \$332 base carries relatively more weight in the middle of the distribution than a pure percentage-of-income tax would. One caveat on the bottom decile specifically: a small number of tax units with reported business losses rank into decile 1 by total PUMS income (`tu_income_dist`, which is loss-inclusive) while still owing a positive 2a liability on their floored wage/interest income — the same negative-income-bin artifact flagged for the property-tax figures above and for Part II's household-ranked income tables; read D1 here as elevated by that artifact, not as a precise rate. Third, neither hand-code captures the federal side of the tax code — the EITC, refundable child credits, or the federal SALT deduction — because ACS PUMS carries no itemization detail and Ohio's schedule has no federal-liability dependency to hand-code against. That gap is exactly why Part II's income-incidence engine (`02_code/3b-income-incidence.R`) makes a separate, narrow TAXSIM call for *federal* marginal rates alone (a documented 2023-law stable-structure proxy) rather than for anything on the Ohio side — the federal interaction motivates that one remaining TAXSIM call, not a TAXSIM run for Ohio.

**Anchor:** for the rest of the analysis, including every scenario in Part II, we treat **2a's hand-coded TY2026 Ohio law** as the income-tax baseline; TAXSIM under 2023 law is used only for federal marginal rates in `3b`. Replacement scenarios solve against this baseline and report the *change* in burden against it (`05_output/tables/winners_losers_by_decile_tenure.csv`, Part II).

## Where the replacement-tax incidence stands now

The baseline built above — property-tax incidence by decile and tenure, and the income-tax construction that closes with 2a's anchored TY2026 schedule — is no longer waiting on data. Three of the four legs the replacement model needs are built. Sales-tax incidence draws on the rebuilt Consumer Expenditure Survey chain (`02_code/1d`-`1g`, `1m`; figures 25-27 below), which turns these income-only PUMS households into a full consumption profile and feeds `02_code/3c-sales-incidence.R`. State income-tax incidence uses 2a's hand-coded TY2026 baseline (above), which feeds `02_code/3b-income-incidence.R`'s household-ranked microsim. Both incidence engines, together with the Layer-1 replacement-rate solver and the winners/losers tables, are stood up in the scenario matrix — see `tax_scenarios_model.qmd` (Part II) for the full model, its sensitivity program, and its reconciliation trail.

Local income tax is the one leg that remains a genuine household-incidence blocker, not a data gap the pipeline can close by pulling another file. Municipal income tax is levied where a person **works**; school-district income tax where they **live, by district**. PUMS resolves only to **PUMA**, which is neither a municipality nor a school district, so neither tax can be assigned to a household without finer geography than the data provide. Local income tax therefore stays an aggregate (collections) story — mapped in section iii above — rather than a household-incidence one, in this model.

# Analyzing CEX data for sales-tax incidence

The Consumer Expenditure Survey (CEX) provides detailed information on household spending through its Diary and Interview surveys. While there is no representative Ohio-specific CEX sample, we use the national CEX and apply Ohio's tax rules through the taxability crosswalk (a Midwest-only robustness check and an Ohio-only audit diagnostic are planned; see `MODEL_OUTLINE_v02_behavioral.md`). The prep layer (rebuilt 2026-07-10) follows the official BLS calendar-year method: all five interview quarters contribute with months-in-scope weights, expenditures are assigned by reference month, and the two surveys are integrated UCC-by-UCC using the BLS integrated hierarchy (each UCC sourced from exactly one survey). The rebuilt means reproduce BLS published category totals within \~2% for every category except 2022 cash contributions (+17%, flagged in the audit for a published-value recheck) (audit: `05_output/audit/recon_cex_rebuild_vs_bls_published.csv`).

The integrated expenditure universe has *675* UCC items to map to the *28* industrial NAICS-based classifications used by the Ohio tax dataset. The manually reviewed UCC-to-NAICS crosswalk (initially ChatGPT-assisted, then hand-checked) lives in the "UCC Mapping" sheet of `diagnostic_ucc_summary_with_naics_mapping.xlsx`; the merge script `1f-merge-ucc-manual.R` carries the manual selections forward whenever the prep layer is rebuilt. The crosswalk maps each UCC to an Ohio sales-tax NAICS bucket and flags UCCs that are non-taxable or not directly allocable to Ohio sales/use collections. Five buckets (notably food & beverage retail and gasoline stations, \~\$182M combined) currently have collections but no mapped CEX spending — an outlet-vs-product classification gap under review (`diagnostic_bucket_coverage.csv`).

```{r}
#| label: prep-industry-total-share
# Read early so the prose immediately below can inline the motor-vehicle
# share live rather than hand-type it; the fig25 chunk below re-reads the
# same file to build the plot.
industry_total_early <- readr::read_csv(here("data/clean/intermediate/cex/sales_tax_burden/plot_data_industry_total_allocated_tax.csv"),
                                        show_col_types = FALSE)
motor_vehicle_share <- industry_total_early$allocated_tax_burden[industry_total_early$bucket_name == "Motor vehicle and parts dealers"] /
  sum(industry_total_early$allocated_tax_burden, na.rm = TRUE)
```

Excluding the non-taxable and non-allocable UCC items, approximately `r scales::percent(motor_vehicle_share, accuracy = 0.1)` of household-borne S&U taxes are from retail trade of motor vehicle and parts dealers, with general merchandise, building material and garden retailers, accommodation and food service, and other miscellaneous retailers making up the next largest shares.

```{r}
#| label: fig-sales-tax-borne-by-households-collected-by-industry
#| fig-cap: "Ohio sales and use tax collections borne by households (excl. business-to-business expenditures), by industry —  (2025 Ohio Tax Dept)."
industry_total <- readr::read_csv(here("data/clean/intermediate/cex/sales_tax_burden/plot_data_industry_total_allocated_tax.csv"),
                                  show_col_types = FALSE)

p_industry <- industry_total |>
  mutate(
    pct_total = allocated_tax_burden / sum(allocated_tax_burden, na.rm = TRUE),
    pct_total_label = scales::percent(pct_total, accuracy = 0.1),
    bucket_name = fct_reorder(bucket_name, allocated_tax_burden, .desc = TRUE)
  ) |>
  ggplot(aes(x = bucket_name, y = allocated_tax_burden)) +
  geom_col(fill = swank_palette[1], width = 0.72) +
  geom_text(aes(label = pct_total_label), vjust = -0.35, size = 3.2, color = "#404040") +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M"), expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Allocated (to consumers) Ohio county sales/use tax burden by industry",
       subtitle = "Total allocable burden across demographic groups",
       x = NULL, y = "Allocated tax burden",
       caption = "Source: data/clean/intermediate/cex/sales_tax_burden/plot_data_industry_total_allocated_tax.csv (02_code/1g).") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
p_industry
ggsave(file.path(fig_dir, "fig25_cex_sales_tax_burden_by_industry.png"),
       p_industry, width = 9, height = 6, dpi = 300, bg = "white")
```

Expenditure on different categories varies by income group. The following figure shows the share of total expenditure on each level 2 category (as defined by the UCC hierarchy) by income group. The shares sum to 100% within each income group.

```{r}
#| label: fig-CEX-expenditure-by-income-group-level-2-UCC
#| fig-cap: "Consumer Expenditures by Category — (2022-2024 Consumer Expenditure Survey)."
# Expenditure-share-by-income-group-and-category prep now lives in
# 02_code/1m-cex-burden-tables.R (promoted from this chunk's former ~150-line
# inline CEX merge -- see 09_quality_reports/plans/2026-07-13_phase3-model-
# descriptives.md, Phase 3C); this chunk only reads and plots.
income_agg_order <- c("<$50k", "$50k-$99,999", "$100k-$149,999", "$150k+")

ce_cu_bucket_spending_demo <- readr::read_csv(
  here("data/clean/intermediate/cex/sales_tax_burden/expenditure_share_by_income_group_level2.csv"),
  show_col_types = FALSE
) |>
  mutate(income_group_agg = factor(income_group_agg, levels = income_agg_order),
        share_label = scales::percent(share_exp, accuracy = 0.1))

p_income_bucket_share_grouped <- ce_cu_bucket_spending_demo |>
  ggplot(aes(x = level_2_label, y = share_exp, fill = income_group_agg)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  geom_text(aes(label = share_label), position = position_dodge(width = 0.8), vjust = -0.25, size = 3) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = swank_palette[1:4], name = NULL) +
  labs(title = "Share of CEX expenditures by level-2 category and income group",
       subtitle = "Shares sum to 100% within each income group",
       x = NULL, y = "Share of income group's total expenditure",
       caption = "Source: data/clean/intermediate/cex/sales_tax_burden/expenditure_share_by_income_group_level2.csv (02_code/1m).") +
  theme_swank() +
  theme(legend.position = "top", panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
p_income_bucket_share_grouped
ggsave(file.path(fig_dir, "fig26_cex_expenditure_share_by_income_group.png"),
       p_income_bucket_share_grouped, width = 10, height = 6, dpi = 300, bg = "white")
```

Because different products and services face different sales-tax rates and exemptions, the incidence of the sales tax varies across income groups. The figure below combines the CEX expenditure data with Ohio's sales-tax rates and exemptions, via the UCC-NAICS crosswalk, to estimate the sales-tax burden by income group directly — the descriptive counterpart to the property-tax incidence figures above, and the input `02_code/3c-sales-incidence.R` carries forward into Part II's replacement-scenario incidence tables.

```{r}
#| label: fig-CEX-OH-sales-tax-burden-by-income-group
#| fig-cap: "Average Sales Tax Burden as a Share of Income — 2025 tax collections based on 2022-2024 CEX."
# Sales-tax-burden-by-income-group prep now lives in 02_code/1m-cex-burden-
# tables.R (promoted from this chunk's former inline aggregation -- see
# 09_quality_reports/plans/2026-07-13_phase3-model-descriptives.md, Phase 3C);
# this chunk only reads and plots.
income_order <- c("<$25k", "$25k-$49,999", "$50k-$74,999", "$75k-$99,999",
                  "$100k-$149,999", "$150k+")

income_tax_burden_2025 <- readr::read_csv(
  here("data/clean/intermediate/cex/sales_tax_burden/sales_tax_burden_by_income_group.csv"),
  show_col_types = FALSE
) |>
  mutate(income_group = factor(income_group, levels = income_order),
        tax_burden_pct_income_label = scales::percent(tax_burden_pct_income, accuracy = 0.01))

p_income_tax_burden_2025 <- income_tax_burden_2025 |>
  ggplot(aes(x = income_group, y = tax_burden_pct_income)) +
  geom_col(fill = swank_palette[1], width = 0.7) +
  geom_text(aes(label = tax_burden_pct_income_label), vjust = -0.35, size = 3.5, color = "#404040") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1), expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Average sales tax burden as a share of income",
       subtitle = "2025 allocated sales/use tax burden divided by average annual CU income, 2022-2024",
       x = NULL, y = "Sales tax burden as share of income",
       caption = "Source: data/clean/intermediate/cex/sales_tax_burden/sales_tax_burden_by_income_group.csv (02_code/1m).") +
  theme_swank() +
  theme(panel.grid.major.x = element_blank())
p_income_tax_burden_2025
ggsave(file.path(fig_dir, "fig27_cex_sales_tax_burden_by_income_group.png"),
       p_income_tax_burden_2025, width = 8, height = 5, dpi = 300, bg = "white")
```

